Raising the role of Vocabulary Hubs for Semantic data interoperability in Dataspaces

### 2024-06-11 EBDVF-2024/paper.qmd
Added introduction and current situation (projects and use cases).

# Introduction
Consolidating wind farm data in data spaces for training predictive maintenance ML - Use Case Description

# Initial Situation
In the UNDERPIN project, we face the challenge of training ML for predictive maintenance of wind farms. This scenario benefits from large amounts of training data. Such data can be provided by different partners who collaborate on this ML approach. We support this scenario by implementing a data space for wind farm data. Data spaces enable data sharing between trusted participants while keeping data sovereignty. However, for consolidating different data sets from participants, we need to establish a clear semantics and mapping for the data.

>>>>

# Introduction
The european data economy heavily depends on the availability of data and the technological foundation to make use of it. Different industries and communities have a mutual benefit of sharing data between each other and thereby support their business goals. Examples are the well-known mobility data space, but also use cases like energy communities for running simulations and machine learning based systems, which heavily rely on high amount of high qualitative training data. To cope with the challenge of data sharing, data space approaches like IDS were introduced to not only solve the technical aspects of data sharing, but also to establish methods for data sovereignty so that data space participants can determine themselves how and when others can make use of the shared data.
For making use of shared data, interoperability is a crucial factor. While syntactic data exchange is well covered by clearly defined data formats, there is still the open challenge of semantic interoperability. IDS supports semantic interoperability by leveraging Semantic Web standards and architectural components, like the Vocabulary Hub, which stores semantic vocabularies for metadata and data descriptions and can dereference URIs to provide semantic details of data assets.
In this paper, we present our proposal for a technological solution of an IDS Vocabulary Hub build on the integration of the GraphDB graph database and PoolParty Semantic Suite. We not only cover the core Vocabulary Hub functionalities and responsibilities in a data space, but also extend the idea towards a Semantic Layer for data spaces, which provides even more services to easily and semi-automatically manage semantic interoperability.

# Projects and Use Cases
In the expanding data economy landscape supported by data spaces, we are working in several projects which address the practical needs and were we experience the importance of semantic interoperability to fulfill various use cases. Here we present 3 examples of from the DataBri-X and UNDERPIN projects, which cover various application domains.

## The Energy Data Space
In this use case, we run simulations for energy communities, which predict the behavior of the energy grid. To provide precise predictions, there is the need of high amounts of example data. This data is of numeric type and different tabular data sources need to be semantically consolidated.

## The Legal Data Space
Analyzing documents in the legal domain is a highly challenging task, because of the complexity of legal information. We provide recommender services for insights into corpora of legal documents, which identify the meaning via a legal knowledge graph, NLP analysis and semantic annotations.

## The Wind Farm Data Space
Operating wind farms involves high costs for maintaining the wind turbines. Predicting potential failures and optimizing the maintenance operations to minimize the costs can be business critical. For establishing predictive maintenance, we need high amounts of high qualitative training data, which can be achieved by consolidation different wind farm measurement sources. Again, we have the case where we need semantic interoperability for numeric tabular data.

These are just 3 examples to illustrate the heterogeneity of data in the context of different use cases. We aim to develop solutions which not only fulfill the data space needs, but also improve the situation regarding semantic interoperability by providing services which can be easily leveraged to implement such use cases.
